---
title: "Effects of Wildfires on Covid"
output: html_document
date: "`r format(Sys.time(), '%d %B, %Y')`"
urlcolor: blue
---

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.pos="!H", out.extra = "")
```

# Independent Variable: Air Quality

[Github repo](github.com/mooper/fires-covid)

## Option 1: Data on PM2.5 from AirNow Sensors.

- Missing in some counties
- Not yet corrected for outliers/faulty stations

![PM2.5 in counties with AirNow Sensors, in August and September 2020.](/home/mattcoop/fires-covid/res/pm25-crop.pdf)


## Option 2: Data on Aerosol Index from ESA Sentinel Satellite.

- More comprehensive
- "The main aerosol types that cause signals detected in the AI are desert dust, biomass burning and volcanic ash plumes." [here](https://earth.esa.int/web/sentinel/technical-guides/sentinel-5p/level-2/aerosol-index)

![Aerosol Index, in August and September 2020.](/home/mattcoop/fires-covid/res/aerosols-crop.pdf)

\newpage

# Dependent Variable: Covid Cases & Mortality

- Data from [USAfacts.com](USAfacts.com), was recommended by CDC.
- Comprehensive data at daily, county-level scale.
- In some cases, states are at the state level:

![Percent of cases and deaths unassigned to a specific county since July.](/home/mattcoop/fires-covid/res/unassigned-crop.pdf)

So, we will model for CA, OR, WA, NV, and ID

\newpage

# Modeling

- Focus on past month
- Fixed effects at the county level
- Thin-Plate Spline temporal effect
- Day-of-week fixed effect

```{r, echo=FALSE}
extract.gam <- function(x){
  tidy(x, parametric=T) %>%
    filter(!grepl('countyFIPS|doy', term))
}

```
```{r}
library(lubridate)
library(tidyverse)
library(broom)

setwd('/home/mattcoop/fires-covid/data/')

options(stringsAsFactors=F)

dat <- read.csv('moddat.csv') %>% 
  filter(State %in% c("CA", "WA", "OR", "ID", "NV"),
         date >= ymd('2020-08-15'),
         date < ymd('2020-09-14')) %>%
  mutate(doy = factor(weekdays(ymd(date))),
         countyFIPS = factor(countyFIPS),
         date = as.numeric(ymd(date)))

death_pm25 <- gam(death_rate ~ pm25 + countyFIPS + s(date) + doy,
                  data = dat, method='REML')
case_pm25 <- gam(case_rate ~ pm25 + countyFIPS + s(date) + doy,
                 data = dat, method='REML')
death_aeros <- gam(death_rate ~ aerosols + countyFIPS + s(date) + doy,
                      data = dat, method='REML')
case_aeros <- gam(case_rate ~ aerosols + countyFIPS + s(date) + doy,
                     data = dat, method='REML')

extract.gam(death_pm25)
extract.gam(case_pm25)
extract.gam(death_aeros)
extract.gam(case_aeros)

```
